Implementing Equational Constraints\\ in a Functional Language

### 2011-07-13 papers/wlp2011/paper.tex
Small corrections to paper

As mentioned in the introduction, occurrences of logic variables are
translated into generators.
For instance, the expression \ccode{not x}, where \code{x} is a logic variable,
is translated into \ccode{not (aBool s)}, where s is an \code{IDSupply}
provided by the context of the expression. The latter expression is 
evaluated by reducing the
argument \code{aBool s} to a choice between \code{True} or \code{False}
followed by applying \code{not} to this choice.
This is similar to a narrowing step on \ccode{not x}
that instantiates the variable \code{x} to \code{True} or \code{False}.
% Actually, it has been shown in \cite{AntoyHanus06ICLP}
% that narrowing expressions with logic variables
% is equivalent to rewriting with generators.
Since such generators are standard
non-deterministic operations, they are translated like any other operation
and therefore do not require any additional run-time support.
However, in the presence of equational constraints,
there are methods which are more efficient than generating all values.
These methods and their implementation are discussed in the
next section.

@@

To avoid these problems, we have to implement the idea
of the well-known unification principle \cite{Robinson65}.
Instead of enumerating all values for logic variables
occurring in an equational constraint,
we \emph{bind} the variables to another variable or term.
Since we compile into a purely functional language,
the binding cannot be performed by some side effect.
Instead, we add binding constraints to the computed results to be 
processed during search.

@@

As mentioned above, the consideration of free variables
is relevant in equational constraints where \emph{binding constraints} are
generated. For this purpose, we introduce a type to represent
a binding constraint as a pair of a choice identifier
and a decision for this identifier:
\begin{haskell}
  data Constraint = ID :=: Decision
\end{haskell}
Furthermore, we extend each data type by the possibility to
add constraints:
\begin{haskell}
  data Bool   = $\ldots$ | Guard [Constraint] Bool
  data List a = $\ldots$ | Guard [Constraint] (List a)
\end{haskell}
Thus, \code{(Guard $cs$ $v$)} represents a \emph{constrained value},
i.e., the value $v$ is only valid if the constraints $cs$ are
consistent with the decisions previously made during search.
These binding constraints are created by the
equational constraint operation \ccode{=:=}: if a free variable
should be bound to a constructor, we make the same
decisions as it would be done in the successful branch
of the generator. In case of Boolean values,
this can be implemented by the following additional rules
for \ccode{=:=}:
\begin{haskell}
  Choice (FreeID i) _ _ =:= True   =  Guard [i :=: ChooseLeft ] Success
  Choice (FreeID i) _ _ =:= False  =  Guard [i :=: ChooseRight] Success
\end{haskell}
Hence, the binding of a variable to some known value
is implemented as a binding constraint for the choice identifier
for this variable. However, if we want to bind a variable
to another variable, we cannot store a concrete decision.
Instead, we store the information that the decisions for
both variables, when they are made to extract values,
must be identical. For this purpose, we extend the \code{Decision}
type to cover this information:
\begin{haskell}
  data Decision = $\ldots$ | BindTo ID
\end{haskell}
Furthermore, we add the rule that an equational constraint
between two variables yields a binding for these variables:
\begin{haskell}
  Choice (FreeID i) _ _ =:= Choice (FreeID j) _ _
    =  Guard (i :=: BindTo j) Success
\end{haskell}
%
The consistency of constraints is checked when values are extracted
from a choice structure, e.g., by the operation \code{printValsDFS}.
For this purpose, we extend the definition of the corresponding search
operations by calling a constraint solver for the constraints.
For instance, the definition of \code{printValsDFS} is extended by
a rule handling constrained values:
%
\begin{haskell}
  $\ldots$
  printValsDFS (Guard cs x) = do consistent <- add c
                                 if consistent then do printValsDFS x
                                                       remove c
                                               else return ()
  $\ldots$
\end{haskell}
The operation \code{add} checks the consistency of the constraint \code{c}
with the decisions made so far and, in case of consistency,
stores the decision made by the constraint.
In this case, the constrained value is evaluated
before the constraint is removed to allow backtracking.
Furthermore, the operations \code{lookupDecision} and \code{setDecision}
are extended to deal with bindings between two variables,
i.e., they follow variable chains in case of \code{BindTo} constructors.

Finally, with the ability to distinguish free variables 
(choices with an identifier of the form \code{(FreeID $\ldots$)})
from other values during search, values containing logic variables 
can also be printed in a specific form rather than enumerating all values, 
like in logic programming systems. For instance, KiCS2 evaluates the 
application of \code{head} to an unknown list as follows:
\begin{haskell}
  Prelude> head xs where xs free
  {xs = (_x2:_x3)} _x2
\end{haskell}
Here, free variables are marked by the prefix \code{\us{}x}.

@@

The general structure of the implementation of functional patterns in KiCS2
is quite similar to that of equational constraints,
with the exception that variables could be also bound
to unevaluated expressions. 
Only if such variables are later
accessed, the expressions they are bound to are evaluated.
This can be achieved by adding a further alternative
to the type of decisions:
\begin{haskell}
  data Decision = $\ldots$ | LazyBind [Constraint]
\end{haskell}
%
The implementation of the lazy unification operation \ccode{=:<=}
is almost identical to the strict unification operation \ccode{=:=}
as shown in Section~\ref{sec:Unification}.
The only difference is in the rules where a free variable occurs
in the left argument. All these rules are replaced by the single
rule
\begin{haskell}
  Choice (FreeID i) _ _ =:<= x
    = Guard [i :=: LazyBind (lazyBind i x)] Success
\end{haskell}
where the auxiliary operation \code{lazyBind}
implements the demand-driven evaluation of the right argument \code{x}:
\begin{haskell}
  lazyBind :: ID -> a -> [Constraint]
  lazyBind i True  = i :=: [ChooseLeft]
  lazyBind i False = i :=: [ChooseRight]
\end{haskell}
The use of the additional \code{LazyBind} constructor allows the argument 
\code{x} to be stored in a binding constraint without evaluation 
(due to the lazy evaluation strategy of the
target language Haskell).
However, it is evaluated by \code{lazyBind} when its binding
is required by another part of the computation.
Similarly to equational constraints,
lazy bindings are processed by a solver when values are extracted.
In particular, if a variable has more than one lazy binding constraint
(which is possible if a functional pattern evaluates to a non-linear term),
the corresponding expressions are evaluated and unified
according to the semantics of functional patterns
\cite{AntoyHanus05LOPSTR}.

@@

In contrast to the Curry implementations PAKCS and MCC, our approach 
of strict unification is based
on an explicit representation of the search space instead of
backtracking and manipulating
a global state containing bindings for logic variables.
Nevertheless, the benchmarks in Fig.~\ref{fig:unification},
using equational constraints only,
show that it can compete with or even outperform the other implementations.
The results show that the implementation of unification
of MCC performs best. However, in most cases our implementation 
outperforms the Prolog-based PAKCS implementation, 
except for some examples where further investigations
and optimizations are necessary.

>>>>

As mentioned in the introduction, occurrences of logic variables are
translated into generators.
For instance, the expression \ccode{not x}, where \code{x} is a logic variable,
is translated into \ccode{not (aBool s)}, where \code{s} is an \code{IDSupply}
provided by the context of the expression. The latter expression is 
evaluated by reducing the
argument \code{aBool s} to a choice between \code{True} or \code{False}
followed by applying \code{not} to this choice.
This is similar to a narrowing step on \ccode{not x}
that instantiates the variable \code{x} to \code{True} or \code{False}.
% Actually, it has been shown in \cite{AntoyHanus06ICLP}
% that narrowing expressions with logic variables
% is equivalent to rewriting with generators.
Since such generators are standard
non-deterministic operations, they are translated like any other operation
and, therefore, do not require any additional run-time support.
However, in the presence of equational constraints,
there are methods which are more efficient than generating all values.
These methods and their implementation are discussed in the
next section.

@@

To avoid these problems, we have to implement the idea
of the well-known unification principle \cite{Robinson65}.
Instead of enumerating all values for logic variables
occurring in an equational constraint,
we \emph{bind} the variables to another variable or term.
Since we compile into a purely functional language,
the binding cannot be performed by some side effect.
Instead, we add binding constraints to the computed results to be 
processed by a search strategy that extract values from choice structures.

@@

As mentioned above, the consideration of free variables
is relevant in equational constraints where \emph{binding constraints} are
generated. For this purpose, we introduce a type to represent
a binding constraint as a pair of a choice identifier
and a decision for this identifier:
\begin{haskell}
  data Constraint = ID :=: Decision
\end{haskell}
Furthermore, we extend each data type by the possibility to
add constraints:
\begin{haskell}
  data Bool   = $\ldots$ | Guard [Constraint] Bool
  data List a = $\ldots$ | Guard [Constraint] (List a)
\end{haskell}
Thus, \code{(Guard $cs$ $v$)} represents a \emph{constrained value},
i.e., the value $v$ is only valid if the constraints $cs$ are
consistent with the decisions previously made during search.
These binding constraints are created by the
equational constraint operation \ccode{=:=}: if a free variable
should be bound to a constructor, we make the same
decisions as it would be done in the successful branch
of the generator. In case of Boolean values,
this can be implemented by the following additional rules
for \ccode{=:=}:
\begin{haskell}
  Choice (FreeID i) _ _ =:= True   =  Guard [i :=: ChooseLeft ] Success
  Choice (FreeID i) _ _ =:= False  =  Guard [i :=: ChooseRight] Success
\end{haskell}
Hence, the binding of a variable to some known value
is implemented as a binding constraint for the choice identifier
for this variable. However, if we want to bind a variable
to another variable, we cannot store a concrete decision.
Instead, we store the information that the decisions for
both variables, when they are made to extract values,
must be identical. For this purpose, we extend the \code{Decision}
type to cover this information:
\begin{haskell}
  data Decision = $\ldots$ | BindTo ID
\end{haskell}
Furthermore, we add the rule that an equational constraint
between two variables yields a binding for these variables:
\begin{haskell}
  Choice (FreeID i) _ _ =:= Choice (FreeID j) _ _
    =  Guard (i :=: BindTo j) Success
\end{haskell}
%
The consistency of constraints is checked when values are extracted
from a choice structure, e.g., by the operation \code{printValsDFS}.
For this purpose, we extend the definition of the corresponding search
operations by calling a constraint solver for the constraints.
For instance, the definition of \code{printValsDFS} is extended by
a rule handling constrained values:
%
\begin{haskell}
  $\ldots$
  printValsDFS (Guard cs x) = do consistent <- add cs
                                 if consistent then do printValsDFS x
                                                       remove cs
                                               else return ()
  $\ldots$
\end{haskell}
The operation \code{add} checks the consistency of the constraint \code{cs}
with the decisions made so far and, in case of consistency,
stores the decision made by the constraint.
In this case, the constrained value is evaluated
before the constraint is removed to allow backtracking.
Furthermore, the operations \code{lookupDecision} and \code{setDecision}
are extended to deal with bindings between two variables,
i.e., they follow variable chains in case of \code{BindTo} constructors.

Finally, with the ability to distinguish free variables 
(choices with an identifier of the form \code{(FreeID $\ldots$)})
from other values during search, values containing logic variables 
can also be printed in a specific form rather than enumerating all values, 
similarly to logic programming systems. For instance, KiCS2 evaluates the 
application of \code{head} to an unknown list as follows:
\begin{haskell}
  Prelude> head xs where xs free
  {xs = (_x2:_x3)} _x2
\end{haskell}
Here, free variables are marked by the prefix \code{\us{}x}.

@@

The general structure of the implementation of functional patterns in KiCS2
is quite similar to that of equational constraints,
with the exception that variables could be also bound
to unevaluated expressions. 
Only if such variables are later
accessed, the expressions they are bound to are evaluated.
This can be achieved by adding a further alternative
to the type of decisions:
\begin{haskell}
  data Decision = $\ldots$ | LazyBind [Constraint]
\end{haskell}
%
The implementation of the lazy unification operation \ccode{=:<=}
is almost identical to the strict unification operation \ccode{=:=}
as shown in Section~\ref{sec:Unification}.
The only difference is in the rules where a free variable occurs
in the left argument. All these rules are replaced by the single
rule
\begin{haskell}
  Choice (FreeID i) _ _ =:<= x
    = Guard [i :=: LazyBind (lazyBind i x)] Success
\end{haskell}
where the auxiliary operation \code{lazyBind}
implements the demand-driven evaluation of the right argument \code{x}:
\begin{haskell}
  lazyBind :: ID -> a -> [Constraint]
  lazyBind i True  = [i :=: ChooseLeft]
  lazyBind i False = [i :=: ChooseRight]
\end{haskell}
The use of the additional \code{LazyBind} constructor allows the argument 
\code{x} to be stored in a binding constraint without evaluation 
(due to the lazy evaluation strategy of the
target language Haskell).
However, it is evaluated by \code{lazyBind} when its binding
is required by another part of the computation.
Similarly to equational constraints,
lazy bindings are processed by a solver when values are extracted.
In particular, if a variable has more than one lazy binding constraint
(which is possible if a functional pattern evaluates to a non-linear term),
the corresponding expressions are evaluated and unified
according to the semantics of functional patterns
\cite{AntoyHanus05LOPSTR}.

@@

In contrast to the Curry implementations PAKCS and MCC,
our implementation of strict unification is based
on an explicit representation of the search space instead of
backtracking and manipulating
a global state containing bindings for logic variables.
Nevertheless, the benchmarks in Fig.~\ref{fig:unification},
using equational constraints only,
show that it can compete with or even outperform the other implementations.
The results show that the implementation of unification
of MCC performs best. However, in most cases our implementation 
outperforms the Prolog-based PAKCS implementation, 
except for some examples where further investigations
and optimizations are necessary.
